Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-31 paper2.qmd
section:4.2; process:write,finish

<!-- 1. What is the more realistic scenario: the sample mechanism -->
In the sample-comparison model, we assume a more realistic scenario than the conceptual-population model. Specifically, we assume researchers have access to a data sample consisting of a *limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$*, drawn from the conceptual population. In a subsequent step, we assume that judges *do not make all possible comparisons* allowed by the sample, but rather only $n_{C}$ comparisons.

@@

<!-- 1. start by defining the sampling vector and matrices  -->
To incorporate the sampling mechanism and facilitate the interpretation of the sample-comparison model, we first define the data sampling process using the binary vector variables $S_{I}$, $S_{J}$, $S_{IA}$, and $S_{JK}$ as follows:

@@

$$
i_{(i)} = \begin{cases} 
1 & \text{if the data for the } i \text{ element of } I \text{ is observed} \\
0 & \text{if the data for the } i \text{ element of } I \text{ is missing}
\end{cases}
$$ {#eq-mat22}

@@

$$
ia_{(ia)} = \begin{cases} 
1 & \text{if the data for the } i,a \text{ elements of } IA \text{ are observed} \\
0 & \text{if the data for the } i,a \text{ elements of } IA \text{ are missing}
\end{cases}
$$ {#eq-mat23}

@@

Finally, we define the *sample design* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$. The sum of its elements represents the total data sample $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$. Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals from the population.

@@

The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$. A square encloses this vector, indicating that it is a conditioned variable. In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp. 32], in this case $S$. Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector determines *which repeated comparisons the subset of judges makes for the subset of stimuli produced by the subset of individuals* sampled from the conceptual-population. This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.

Notably, the DAG shows that $S$ is independent of all other variables in the model. This implies that the conceptual model represented in [DAG -@fig-cj12_dag] applies exclusively to Simple Random Sampling (SRS) designs [@Kohler_et_al_2019, pp. 150]. In these designs, each judgment, judge, stimulus, and individual has the same probability of being included in the sample as any other observation within their respective groups [@Lawson_2015, pp. 16-112].

@@

However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals --due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p. 5]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.

@@

<!-- 1. What is the comparison mechanism -->
<!-- Conversely, we make the scenario depicted in @sec-theory-theoretical_SC1 more realistic by integrating the comparison mechanism, that is, we assume that researchers do not have access to *all comparisons* from the sample, but rather a sufficient number of comparisons $(n_{C})$ so that a proportion, $P(B>A)$, may be determined for each pair of stimuli [@Thurstone_1927b, pp. 267]. @fig-cj14 illustrates this model, incorporating both the sample and comparison mechanisms.  -->

@@

@@

<!-- Consider algorithm used, as a child of the traits of judges and stimuli. It allows to talk about independence assumption between them and the perception of difference. Consider algorithm is a categorical variable. -->

<!-- 2. What is the effect of the sampling mechanism -->
<!-- Following the convention of @McElreath_2020 [pp. 499-516], the [DAG -@fig-cj14_dag] depicts the sample-comparison outcome $O^{sc}_{R}$ as an unobserved variable, emphasizing the researchers' inability to access such an outcome due to the comparison mechanism. Moreover, the DAG shows the conditioned *comparison design* variable $C$, which causally influence the observed outcome $O_{R}$. This variable reflects how this observed outcome is a sample drawn from the sample-comparison outcome $O^{sp}_{R}$. Specifically, the variable is a vector $C = \{ c_{1}, \dots, c_{n} \}$, where $n$ corresponds to the number of rows in the matrix $R$ (refer to @sec-theory-theoretical_P2), and each $c_{i}$ is a binary variable indicating the presence $(1)$ or absence $(0)$ of data in the corresponding positions of the sample-comparison outcome $O^{sp}_{R}$. Specifically, -->

$$
c_i = \begin{cases} 
1 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{R} \text{ is observed} \\
0 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{R} \text{ is missing}
\end{cases}
$$

@@

when you talk about the variability, refer that the equal dispersion is just assuming 1 for identification.

@@

::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
  O^{cp}_{R} & := f_{O}(D_{R}) \\
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & := f_{e}(X_{I}) \\
  e_{J} & := f_{e}(Z_{J}) \\
  e_{IA} & \dsep e_{JK}
\end{aligned}
$$

@@

@@

<!-- Thus Thurstone’s model (in both its Case 5 and Rasch formulation) achieves the goal of sample-free calibration (Wright, 1977) in the sense that the estimate of the distance between A and B does not depend on which other objects they are compared with. [@Bramley_2008, pp. 256] -->

@@

>>>>

<!-- 1. What is the more realistic scenario: the sample mechanism -->
The sample-comparison model presents a more realistic scenario than the conceptual-population model. Expressly, we assume researchers work with a data sample consisting of a *limited number of judgments $(n^{s}_{K})$ from a sample of judges $(n^{s}_{J})$ and a specific number of texts $(n^{s}_{A})$ from a sample of students $(n^{s}_{I})$*, all drawn from the conceptual population. In a subsequent step, we assume that judges do not perform *all possible repeated comparisons* within the data sample. Instead, they conduct a sufficient number of comparisons, $n_{C}$, to ensure an accurate estimation of the proportion $P(B>A)$.

@@

<!-- 1. start by defining the sampling vector and matrices  -->
To incorporate the sampling mechanism and facilitate the interpretation of the sample-comparison model, we first define the *data sampling process* using the binary vector variables $S_{I}$, $S_{J}$, $S_{IA}$, and $S_{JK}$ as follows:

@@

$$
i_{(i)} = \begin{cases} 
1 & \text{if data element } i \text{ from } I \text{ is observed} \\
0 & \text{if data element } i \text{ from } I \text{ is missing}
\end{cases}
$$ {#eq-mat22}

@@

$$
ia_{(ia)} = \begin{cases} 
1 & \text{if data elements } i,a \text{ from } IA \text{ are observed} \\
0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
\end{cases}
$$ {#eq-mat23}

@@

Finally, we define the *sample design* vector $S$ in equation ([-@eq-mat25]), which maps each element of $S_{IA}$ to every element of $S_{JK}$. Thus, the vector contains $n$ elements, matching the number of rows in the $R$ matrix, and the sum of its elements represents the total data sample: $n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}$. Here, the term ${n^{s}_{I} \cdot n^{s}_{A} \choose 2}$ represents the binomial coefficient, which quantifies the total number of unique comparisons possible between every pair of sampled stimuli generated by each pair of sampled individuals.

@@

The DAG also depicts the *sample design* vector $S$ as a causal factor influencing the sample-comparison outcome $O^{sc}_{R}$. A square encloses this vector, indicating that it is a conditioned variable. In this context, *conditioning* means restricting the statistical analysis to a specific subset of the population based on the values of a conditioned variable [@Neal_2020, pp. 32], in this case $S$. Essentially, $S$ serves as the sample equivalent for the $R$ matrix (see equation ([-@eq-mat12])), that is, this vector selects *all repeated comparisons a subset of judges makes for the subset of stimuli produced by the sampled individuals*. This process ultimately generates the sample-comparison outcome $O^{sc}_{R}$ from the conceptual-population outcome $O^{cp}_{R}$.

Notably, the DAG shows that $S$ is independent of all other variables in the model. This implies that the conceptual model represented in [DAG -@fig-cj12_dag] applies exclusively to Simple Random Sampling (SRS) designs [@Kohler_et_al_2019, pp. 150]. In these designs, each repeated judgment, judge, stimulus, and individual has the same probability of being included in the sample as any other observation within their respective groups [@Lawson_2015, pp. 16-112].

@@

<!-- 3. segway for next section -->
However, because CJ experiments rarely use the exhaustive pairing of sampled judges, stimuli, and individuals --due to concerns about the practical feasibility of the comparison task, like in @Boonen_et_al_2020 [p. 5]-- a realistic scenario must also account for judges comparing only specific stimuli from certain individuals.

@@

<!-- 1. start by defining the comparison vector  -->
As in the previous section, we first define the *comparison design* using the binary vector variable $C$ to facilitate the interpretation of the sample-comparison model and integrate the comparison mechanism.

$$
C = \begin{bmatrix}
c_{(1)} \\
\vdots \\
c_{(c)} \\
\vdots \\
c_{(n)} \\
\end{bmatrix}
$$ {#eq-mat26}

Where each element of $C$ is a binary value indicating the presence or absence of data rows in the $R$ matrix (see equation ([-@eq-mat12])):

$$
c_{c} = \begin{cases} 
1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
0 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are missing}
\end{cases}
$$

Thus, the *comparison design* vector $C$ contains $n$ elements corresponding to the number of rows in the $R$ matrix, and the sum of its elements represents the selected comparisons, $n_{C}$. This definition assumes that judges do not perform *all possible repeated comparisons* within the data sample. Instead, they conduct a sufficient number of comparisons, $n_{C}$, to ensure the accurate estimation of the proportion $P(B>A)$ for each pair of stimuli [@Thurstone_1927b, pp. 267].

@@

<!-- 2. What is the effect of the comparison mechanism -->
With the definition of $C$, the [DAG -@fig-cj14_dag] incorporates the sample and comparison mechanism into the conceptual-population model. The DAG now also represents the sample-comparison outcome $O^{sc}_{R}$ as unobserved, highlighting that researchers cannot directly access this outcome due to the comparison mechanism. The DAG also represents the *comparison design* vector $C$ as a causal factor influencing the final comparative outcome $O_{R}$. Equivalent to the sampling mechanism, a square encloses this vector, indicating that $C$ determines *which repeated comparisons the subset of judges makes for the subset of stimuli produced by the sampled individuals*. This process ultimately derives the final comparative outcome $O_{R}$ from the sample-comparison outcome $O^{sc}_{R}$.

Notably, the DAG shows that $C$ is also independent of all other variables in the model. This independence implies that the conceptual model represented by [DAG -@fig-cj14_dag] applies exclusively to Random Allocation Comparative Designs [@Bramley_2015, pp. 8] or Incomplete Block Designs [@Lawson_2015, pp. 259-271]. In these designs, every repeated judgment has an equal probability of being included in the sample.

@@

In summary, extending Thurstone's general form with causal inference allows researchers to address BTL modeling issues, such as judge biases (see @sec-theory-issue1b), the hierarchical structure of stimuli (see @sec-theory-issue1b), and measurement errors in trait estimation and hypothesis testing (see @sec-theory-issue2). However, the model does not address concerns about the assumption of equal dispersion discussed in @sec-theory-issue1a. Since this issue involves statistical assumptions regarding the discriminal process distribution, it requires a statistical model, which we will develop in the next section.

@@

<!-- when you talk about the variability, refer that the equal dispersion is just assuming 1 for identification. -->

@@

::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
  O^{cp}_{R} & := f_{O}(D_{R}) \\
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & := f_{e}(X_{I}) \\
  e_{J} & := f_{e}(Z_{J}) \\
  e_{IA} & \dsep e_{JK} \\ 
  \\ 
\end{aligned}
$$

@@

<!-- In order to make it possible to estimate the parameters satisfactorily with the relatively small amount of data available it would probably be necessary to reduce the number of parameters estimated in some other way, for example, by representing all the scripts from one board with a single parameter, as in the logistic regression model used in Greatorex et al. (2002). Such an approach shifts the philosophy of the exercise from that of Thurstone and Rasch (constructing psychological scales capable of yielding sample-free measurement) to the philosophy of statistical modelling (finding a model that optimises parsimony and variance explained). [@Bramley_2008, pp. 274] -->
<!-- Comment: the sample-free measurement philosophy is an illusory one, because in this contexts, the measurements depend on the perception of judges.  -->

@@

@@

<!-- The outcomes of the analysis can be considered in two parts: the ‘model’ and the ‘misfit’. [@Bramley_2008, pp. 260] -->

<!-- It has a good reporting structure, and the pattern of both judge and script fit and misfit enables further investigation of the biases which may be taking place. [@Pollitt_et_al_2003, pp. 7]. -->

<!-- design matters -->

<!-- Thus Thurstone’s model (in both its Case 5 and Rasch formulation) achieves the goal of sample-free calibration (Wright, 1977) in the sense that the estimate of the distance between A and B does not depend on which other objects they are compared with. The practical importance of this is that it is not necessary to have a completely crossed or balanced design of paired comparisons because the estimation process naturally handles missing data. [@Bramley_2008, pp. 256] -->

<!-- if the data fit the Case 5 Thurstone model, the estimate of the separation between any two scripts does not depend on which other scripts they are compared with. This means that data can be missing in a non-random way without affecting the results. However, the precision (standard error) of each script’s estimate depends largely on how many comparisons it has been involved in, so some effort needs to be made to ensure that each script is compared a similar number of times in total. [@Bramley_2008, pp. 260] -->

<!-- Since the analysis can cope with non-random missing data (see earlier), the collection of data can be very flexible – tailored to the needs of the particular situation. Because the comparative judgements are all independent, data can be accumulated as needed. The investigator is thus not (in principle) bound by a prior design. At any point more data could be collected and added to the existing data to recover from accidents, to replace an unacceptable judge, to add extra scripts, to resolve conflicts, to investigate possible bias or to deepen the study of any aspect that proves especially interesting. [@Bramley_2008, pp. 261] -->
